A scenario for a Planckian standard kilogram

### 2017-10-31 manuscript_gravitationalconstant.tex
Final version for first submission

The reform of SI units to use quantum standards is an ongoing ``news
story'' in physics.  At the heart of the reform are two quantum
phenomena that appear on macroscopic scales: the Josephson effect
makes $h/e$ measurable as a voltage-to-frequency ratio, while the
quantum Hall effect makes $h/e^2$ a standard resistance (the
von~Klitzing constant).  These enable electrical units to be defined
in terms of $h$ and $e$.  A further ingenious instrument, known as a
watt balance, precisely compares electrical and mechanical power, thus
enabled mass to be standardized in terms of $h$.  For a simple
introduction to the new SI, we refer the reader to the article on the
``LEGO watt balance''.\cite{Chao15}
The proposed redefinition of the kilogram is an explicit value for
$h$ and $c$, but we can think of it as
\begin{equation}
\mathrm{kg} \equiv 1.3563925\ldots \times 10^{50} \, \mathrm{s}^{-1}
                   \times \frac{h}{c^2}
\end{equation}
with the exact value to be decided.  As we can see, the kilogram is
coupled to the definition of a second.  That is not a problem in
practice, as the second is standardized orders of magnitude better
than any other unit.  Nonetheless, it is interesting to consider a
logical further step of decoupling the kilogram from other
definitions.  This could be done by introducing the gravitational
constant, since $\sqrt{hc/G}$ has dimensions of mass.
Historically, there have been several proposed unit systems that
included $G$ in some way,\cite{Tomilin1999} but the idea of including
both gravity and quantum mechanics in a unit system comes from
Planck.\cite{Planck99} In the process of working towards the radiation
law, Planck realized that $c,G,h$ and the Boltzmann constant $k_B$ (in
his notation $G\rightarrow f$, $h\rightarrow b$, $h/k_B\rightarrow a$)
naturally define units for mass, length, time, and temperature.
Indeed, Planck remarked that such units would be meaningful even in
non-human and extraterrestrial cultures.

A hypothetical Planckian SI unit system would have
\begin{equation} \label{eq:defs}
\begin{aligned}
\mathrm{kg} &= 4.594\ldots \times 10^7
               \left(\frac{\hbar c}{G}\right)^{\!1/2} \\
\mathrm{m}  &= 6.187\ldots \times 10^{34}
               \left(\frac{\hbar G}{c^3}\right)^{\!1/2} \\
\mathrm{s}  &= 1.854\ldots \times 10^{43}
               \left(\frac{\hbar G}{c^5}\right)^{\!1/2} \\
\mathrm{K}  &= 7.058\ldots \times 10^{-33}
               \left(\frac{\hbar c^5}{Gk_B^2}\right)^{\!1/2}
\end{aligned}
\end{equation}
with the numerical coefficients chosen so as to match the present
standards.  The square-rooted expressions are conventionally known as
the Planck mass, Planck length, and so on.  (They are actually
$\sqrt{2\pi}$ smaller than Planck's definitions.)  Extension to
electrical units is straightforward, but we will not discuss it here.
The Planck length, time, and temperature in themselves are extreme
scales, not probed by any known physical process, and presumably the
regime of quantum gravity.  But extreme scales do not harm if they are
only part of some expression, as they are here.  The Planck mass is
small but microscopically measurable: $\sim20\mu\mathrm{g}$.  This is
of the same order as the uncertainty in the prototype kilogram itself.
Hence the kilogram could be redefined as an integer multiple of the
Planck mass, without significantly perturbing its value.

@@

\section{A new approach to measure $G$}
In Fig.~\ref{fig:Gmeasurements} we plot the best-known recent
measurements of the gravitational constant. Three of them
(BIPM-14,\cite{Quinn14} UCI-14,\cite{Newman14}, and
HUST-09\cite{Luo09}) were carried out using a torsion balance, whereas
UZur-06\cite{Schlamminger06} used a beam balance, and
JILA-10\cite{Parks10} measured the separation of two simple pendulums
in a gravitational field. They all tried to measure the very weak
gravitational force with a mechanical apparatus. The sixth measurement
(LENS-14\cite{Rosi14}) was performed with another approach, using cold
atom interferometry. Its uncertainty is a lot larger than the others,
but it could become more competitive in the future. However, a brief
glance on the plotted measurements shows the disagreement of current
$G$-values and suggests a demand for new, potentially different
techniques.

@@

\textbf{Radiation pressure:} The envelope of the cavity needs to protect the two body-system from the constant bombardment of radiation and charged particles from the Sun. Since the activity of the Sun varies much, the spacecraft must be able to correct its position around the freely falling masses inside the cavity.\\
Furthermore, there is thermal radiation inside the cavity, that could disturb the orbit.
Using the Stefan-Boltzmann law, we can relate radiation pressure to the temperature inside the cavity. If we then specify to balls of diameter $d$, we can express the force on the balls as 
\begin{equation}
F = \frac{2 \pi^6 d^2 {k_\mathrm{B}}^4 T^4}{45 c^3 h^3}.
\end{equation}
As we can see, this force is highly temperature dependent. In order to minimize this influence, one should deploy a temperature as small as possible in the cavity. In fact, if a cooling system based on liquid helium is installed, temperature can be brought down to about 4 K. Such a low temperature results in a force on the balls in the order of $10^{-16}$ N. Compared to the gravitational force, this is orders of magnitude less and thus we don't have to worry about effects of radiation pressure.

\textbf{Influence of the position measurement:} 
The photons of the laser that measures the position of the two balls exert an additional force. If a laser with a power of 1 mW is used and not more than five measurements every second are performed, the force on the balls is in the same order as the influence of thermal radiation and thus negligible too.

\textbf{Coulomb force:}
Since the electromagnetic force is orders of magnitude stronger than the gravitational force, the balls must be absolutely neutral. A possible source of charges are cosmic rays or the solar wind, so the two bodies have to be adequately shielded by the enclosing shell.  If we want that the Coulomb force between the balls is in the same order as the effect of thermal radiation and thus negligible, the maximal allowed charge on the balls is about $6 \cdot 10^{-14}\,\mathrm{C}$, which corresponds to roughly $4 \cdot 10^5$ elementary charges.

@@

We now briefly explain in what range the values of the parameters are
estimated.  Numbers are given in Table~\ref{tab:parameters}.  For the
parameter $k$, which is the product of the gravitational constant $G$
and the total mass $M$ we look at the current measured values for the
gravitational constant in Fig.~\ref{fig:Gmeasurements}, it is
reasonable to constrain its numerical value in SI units to values
between $6.671 \cdot 10^{-11}$ and $6.676 \cdot 10^{-11}$. The masses
of the two balls will not contribute to a larger uncertainty in $k$,
we choose them to be 6~kg.  The vector $\vec{\omega}$ accounts for the
rotation of the reference frame. If we say that the orbit of the balls
lies in the same plane as the orbital plane of the Earth,
$\vec{\omega}$ will point in $z$-direction with $x$ and $y$
components small. Thus $\omega_z$ is associated with the period
$P_\odot$ of one year through $2\pi/P_\odot$. Because the rotation
axis are conceivably not exactly aligned, we add some uncertainty. As
mentioned before, the $x$ and $y$ components are small, that is to
say, near zero in roughly in the range of the uncertainty in
$\omega_z$.

@@

In the next step, we set the parameters to a fixed ``true" value
within the range estimated. We have to keep in mind that $k$, $x_0$
and $v_{y_0}$ are linked together and there must not be contradictions
in the true values. The chosen values are displayed in the right
column in Table~\ref{tab:parameters}.  To simulate data we would
obtain from the position measurements, we integrate the equations of
motion. The time for one orbit without perturbations can be estimated
using Kepler's third law (\ref{eq:kepler3}). For the value of $k$
assumed above and a semi-major axis of $0.15$~m, this
results in an orbital period of roughly 12900 seconds or
3.5 hours. Here we integrate two orbits in 100000
steps, that would correspond to about 4 measurements per second,
which should be feasible. To the generated data, we add Gaussian noise
with a standard deviation of 10~nm. The value we use is just an estimate,
it would depend strongly on the measurement system that has been used.

@@

We are not supposing the measurement of the gravitational constant 
using an artificial two-body system at an Earth-Sun Lagrange point to be a stand alone mission. It could just be part of 
a spacecraft reaching a Lagrange point. But our estimates suggest that this approach to increase the precision of the gravitational constant should be taken seriously.

\begin{thebibliography}{99}

>>>>

The reform of SI units to use quantum standards is an ongoing ``news
story'' in physics.\cite{BIPM16}  At the heart of the reform are two quantum
phenomena that appear on macroscopic scales: the Josephson effect
makes $h/e$ measurable as a voltage-to-frequency ratio, while the
quantum Hall effect makes $h/e^2$ a standard resistance (the
von~Klitzing constant).  These enable electrical units to be defined
in terms of $h$ and $e$.  A further ingenious instrument, known as a
watt balance, precisely compares electrical and mechanical power, thus
enabled mass to be standardized in terms of $h$.  For a simple
introduction to the new SI, we refer the reader to the article on the
``LEGO watt balance''.\cite{Chao15}
The proposed redefinition of the kilogram is an explicit value for
$h$ and $c$, but we can think of it as
\begin{equation}
\mathrm{kg} \equiv 1.3563925\ldots \times 10^{50} \, \mathrm{s}^{-1}
                   \times \frac{h}{c^2}
\end{equation}
with the exact value to be decided.  As we can see, the kilogram is
coupled to the definition of a second.  That is not a problem in
practice, as the second is standardized orders of magnitude better
than any other unit.  Nonetheless, it is interesting to consider a
logical further step of decoupling the kilogram from other
definitions.  This could be done by introducing the gravitational
constant, since $\sqrt{hc/G}$ has dimensions of mass.
Historically, there have been several proposed unit systems that
included $G$ in some way,\cite{Tomilin1999} but the idea of including
both gravity and quantum mechanics in a unit system comes from
Planck.\cite{Planck99} In the process of working towards the radiation
law, Planck realized that $c,G,h$ and the Boltzmann constant $k_B$ (in
his notation $G\rightarrow f$, $h\rightarrow b$, $h/k_B\rightarrow a$)
naturally define units for mass, length, time, and temperature.
Indeed, Planck remarked that such units would be meaningful even in
non-human and extraterrestrial cultures.

A hypothetical Planckian SI unit system would have
\begin{equation} \label{eq:defs}
\begin{aligned}
\mathrm{kg} &= 4.594\ldots \times 10^7
               \left(\frac{\hbar c}{G}\right)^{\!1/2} \\
\mathrm{m}  &= 6.187\ldots \times 10^{34}
               \left(\frac{\hbar G}{c^3}\right)^{\!1/2} \\
\mathrm{s}  &= 1.854\ldots \times 10^{43}
               \left(\frac{\hbar G}{c^5}\right)^{\!1/2} \\
\mathrm{K}  &= 7.058\ldots \times 10^{-33}
               \left(\frac{\hbar c^5}{Gk_B^2}\right)^{\!1/2}
\end{aligned}
\end{equation}
with the numerical coefficients chosen so as to match the present
standards.  The square-rooted expressions are conventionally known as
the Planck mass, Planck length, and so on.  (They are actually
$\sqrt{2\pi}$ smaller than Planck's definitions.)  Extension to
electrical units is straightforward, but we will not discuss it here.
The Planck length, time, and temperature in themselves are extreme
scales, not probed by any known physical process, and presumably the
regime of quantum gravity.  But extreme scales do not harm if they are
only part of some expression, as they are here.  The Planck mass is
small but microscopically measurable: $\sim20\mu\mathrm{g}$.  This is
of the same order as the uncertainty in the prototype kilogram itself.\cite{BIPM}
Hence the kilogram could be redefined as an integer multiple of the
Planck mass, without significantly perturbing its value.

@@

\section{A new approach to measure $G$}
In Fig.~\ref{fig:Gmeasurements} we plot the best-known recent
measurements of the gravitational constant. Three of them
(BIPM-14,\cite{Quinn14} UCI-14,\cite{Newman14} and
HUST-09\cite{Luo09}) were carried out using a torsion balance, whereas
UZur-06\cite{Schlamminger06} used a beam balance, and
JILA-10\cite{Parks10} measured the separation of two simple pendulums
in a gravitational field. They all tried to measure the very weak
gravitational force with a mechanical apparatus. The sixth measurement
(LENS-14\cite{Rosi14}) was performed with another approach, using cold
atom interferometry. Its uncertainty is a lot larger than the others,
but it could become more competitive in the future. However, a brief
glance on the plotted measurements shows the disagreement of current
$G$-values and suggests a demand for new, potentially different
techniques.

@@

\textbf{Radiation pressure:} The envelope of the cavity needs to protect the two body-system from the constant bombardment of radiation and charged particles from the Sun. Since the activity of the Sun varies much, the spacecraft must be able to correct its position around the freely falling masses inside the cavity.\\
Furthermore, there is thermal radiation inside the cavity, that could disturb the orbit.
Using the Stefan-Boltzmann law, we can relate radiation pressure to the temperature inside the cavity. If we then specify to balls of diameter $d$, we can express the force on the balls as 
\begin{equation}
F = \frac{2 \pi^6 d^2 {k_\mathrm{B}}^4 T^4}{45 c^3 h^3}.
\end{equation}
As we can see, this force is highly temperature dependent. In order to minimize this influence, one should deploy a temperature as small as possible in the cavity. In fact, if a cooling system based on liquid helium is installed, temperature can be brought down to about 4 K. Such a low temperature results in a force on the balls in the order of $10^{-16}$~N. Compared to the gravitational force, this is orders of magnitude less and thus we don't have to worry about effects of radiation pressure.

\textbf{Influence of the position measurement:} 
The photons of the laser that measures the position of the two balls exert an additional force. If a laser with a power of 1~mW is used and not more than five measurements every second are performed, the force on the balls is in the same order as the influence of thermal radiation and thus negligible too.

\textbf{Coulomb force:}
Since the electromagnetic force is orders of magnitude stronger than the gravitational force, the balls must be absolutely neutral. A possible source of charges are cosmic rays or the solar wind, so the two bodies have to be adequately shielded by the enclosing shell.  If we want that the Coulomb force between the balls is in the same order as the effect of thermal radiation and thus negligible, the maximal allowed charge on the balls is about $6 \cdot 10^{-14}$~C, which corresponds to roughly $4 \cdot 10^5$ elementary charges.

@@

We now briefly explain in what range the values of the parameters are
estimated.  Numbers are given in Table~\ref{tab:parameters}.  For the
parameter $k$, which is the product of the gravitational constant $G$
and the total mass $M$ we look at the current measured values for the
gravitational constant in Fig.~\ref{fig:Gmeasurements}, it is
reasonable to constrain its numerical value in SI units to values
between $6.671 \cdot 10^{-11}$ and $6.676 \cdot 10^{-11}$. The masses
of the two balls will not contribute to a larger uncertainty in $k$,
we choose them to be 6~kg.  The vector $\vec{\omega}$ accounts for the
rotation of the reference frame. If we say that the orbit of the balls
lies in the same plane as the orbital plane of the Earth,
$\vec{\omega}$ will point in $z$-direction with $x$ and $y$
components small. Thus $\omega_z$ is associated with the period
$P_\odot$ of one year through $2\pi/P_\odot$. Because the rotation
axis are conceivably not exactly aligned, we add some uncertainty. As
mentioned before, the $x$ and $y$ components are small, that is to
say, near zero in roughly the range of the uncertainty in
$\omega_z$.

@@

In the next step, we set the parameters to a fixed ``true" value
within the range estimated. We have to keep in mind that $k$, $x_0$
and $v_{y_0}$ are linked together and there must not be contradictions
in the true values. The chosen values are displayed in the right
column of Table~\ref{tab:parameters}.  To simulate data we would
obtain from the position measurements, we integrate the equations of
motion. The time for one orbit without perturbations can be estimated
using Kepler's third law (\ref{eq:kepler3}). For the value of $k$
assumed above and a semi-major axis of $0.15$~m, this
results in an orbital period of roughly 12900 seconds or
3.5 hours. Here we integrate two orbits in 100000
steps, that would correspond to about 4 measurements per second,
which should be feasible. To the generated data, we add Gaussian noise
with a standard deviation of 10~nm. The value we use is just an estimate,
it would depend strongly on the measurement system that has been used.

@@

We are not supposing the measurement of the gravitational constant 
using an artificial two-body system at an Earth-Sun Lagrange point to be a stand alone mission. It could be part of 
a spacecraft reaching a Lagrange point. But our estimates suggest that this approach to increase the precision of the gravitational constant should be taken seriously.

\begin{thebibliography}{99}
\bibitem{BIPM16} International Bureau of Weights and Measures, ``DRAFT 9th edition of the SI Brochure", BIPM, (2016)
